A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period

### 2021-01-05 analysis/paper/paper.Rmd
source 002-005 script files in rmd and cache the data

image_width_1_col_inch = 90/25.4 # image Single column	w = 90 mm 
image_width_2_col_inch = 190/25.4 # image Double column w = 190 mm 
```

@@

# Burial data from Kiwulan, Northeastern Taiwan

@@

<!--archaeological setting-->
We analyzed burial data collected from the upper component of Kiwulan, an Iron Age settlement (1350-1850 AD) located in northeastern Taiwan, which experienced European colonial impacts in the early 17th century and a large wave of Chinese immigrants in the 19th century [@Chen2007]. Excavations revealed abundant pottery sherds, imported ceramics and stonewares, wooden artifacts, stone tools, metal artifacts, imported glass beads and agates beads, and pipes [@Chen2007]. In addition to these artifacts, 90 burials, hundreds of middens, storage pits, and postholes with _in-situ_ posts were also excavated. The burials are mostly located in the middle section of the excavation area, which is the largest open area at Kiwulan that provides continuous stratigraphic sections suitable for temporal comparison (Figure \@ref(fig:burial-map)). Burials are oriented in an east-west direction on the north side of the residential area, indicated by post-holes and `in-situ` wooden posts, indicating a well organized spatial arrangement. Previous studies report an uneven distribution of prestige goods across burials without agreement about whether this uneven distribution hints at vertical social differences. For example, @Cheng2008 interpreted the unequal distribution of glass beads, especially the gold-foil beads between burials, as evidence for hierarchy, indicating a stratified society. However, @Hsieh2012 suggested a relatively egalitarian structure based on comparative analysis of the frequencies of all burial goods. She found that the burials with high value burial goods were usually associated with elders, which might indicate achieved, rather than inherited, status. One important limitation of these previous studies is that they did not use analytical units suitable for comparing pre-European social organization with post-European social organization. Here, we adopt a new chronological framework for the burials to test if network configurations differ from the pre-European period to the post-European period. The discord over the discussion of Kiwulan social organization could be associated with chronological differences that has not yet been well studied.

```{r, burial-map, fig.cap = "Map illustrating the location of burials at Kiwulan (suare size is 4 X 4 meters)", fig.width = image_width_2_col_inch}
knitr::include_graphics(here::here("analysis", "figures", "burial-map.png"))
```

@@

## Data and Hypothesis

<!-- burial data and the definition of tie formation  --> 
We analyzed burial data collected from the published excavation reports, and the original fieldwork notes of Kiwulan. Trade beads are commonly found across burials with substantial differences in quantities, so we described each burial as having one of four levels, high, upper-middle, lower-middle, and low, according to their distributions across all burials. Gold-foil beads are in levels of high (> 3), upper-middle (= 3), lower-middle (= 2), and low (= 1); carnelian beads are in levels of high (> 6), upper-middle (= 4-6), lower-middle (= 3), and low (= 1-2); glass beads are in levels of high (> 6), upper-middle (= 3-6), lower-middle (= 2), and low (= 1). That means if burial 1 and burial 2 both have high quantities of agate beads, then there will be a tie connecting them. For less frequent prestige goods, including imported porcelains, gold foils, and fish-shaped ornaments, we linked two burials when they possess each type of goods in common (i.e. presence or absence). Node attributes here include osteological data, such as age and sex, and cultural data, such as ritual pottery, and burial value index. Ritual pottery was identified as locally made ceramics placed above graves that suggests funeral feasting according to historical records. The burial value index is an important attribute for economical inequality that we assigned the burials into three levels, high (>= 29, account for 10%), above average (= 12-28), and below average (< 12), as an index of wealth. The number is the sum of values of each type of prestige good, which is calculate by the total number of a prestige item from burial context over the number of prestige item in a burial. Since burials tend to have multiple prestige goods in common, the network ties are weighted instead of binary data (where the value 1 represents a tie and the value 0 otherwise) [@Snijders2011]. For example, if two burials have both low quantity of glass beads and porcelain in common, the tie is given a value of 2. Our networks are non-directed, which means ties have no orientation forming the relationship between actors to indicate a mutual relationship.

@@

@@

```{r BERGMs}
```

<!--posterior stats for pre-E-->
We compared estimated statistics from the posterior distributions to examine their differences in structure of the simulated networks (Table XX). For the pre-European network simulations, two nodal covariates, age-homophily and ritual-homophily, have significant effects on the formation of relations between burials. The age homophily shows a negative effect that indicates burials of the same age tend to not share similar prestige goods, while the ritual-homophily demonstrates the positive effect that means burials with ritual pottery tend to form relations. Despite positive values for some covariates, such as sex-homophily and burial value-homophily, they do not show a significant tendency to form relations duo to value of zero in the confidence intervals. Similarly, the dyadic covariate, physical distance, shows positive effects but not significantly, indicating that physical proximity does not affect relations. For the endogenous network effects, transitivity (gwesp) present significant positive effect, while popularity (gwdegree) demonstrates negative effect. The positive value for transitivity suggests a tendency of burials to be clustered in closed transitive structures, indicative of the presence of multiple corporate groups sharing burial goods in common. The negative popularity shows there is a tendency toward decentralization that reflects most burials have similar number of ties without any prominent ones. This might imply that individuals have equal access to trade goods in terms of the flow of goods.

>>>>

image_width_1_col_inch = 90/25.4 # image Single column	w = 90 mm 
image_width_2_col_inch = 190/25.4 # image Double column w = 190 mm 
image_width_3_col_inch = 190/20
```

@@

# Materials

@@

<!--archaeological setting-->
We analyzed burial data collected from the published excavation reports, and the original fieldwork notes for the upper component of Kiwulan, an Iron Age settlement (1350-1850 AD). The settlement experienced European colonial impacts in the early 17th century and a large wave of Chinese immigrants in the 19th century [@Chen2007]. Excavations revealed abundant pottery sherds, imported ceramics and stonewares, wooden artifacts, stone tools, metal artifacts, imported glass beads and agates beads, and pipes [@Chen2007]. In addition to these artifacts, 90 burials, hundreds of middens, storage pits, and postholes with _in-situ_ posts were also excavated. The burials are mostly located in the middle section of the excavation area, which is the largest open area at Kiwulan that provides continuous stratigraphic sections suitable for temporal comparison (Figure \@ref(fig:burial-map)). Burials are oriented in an east-west direction on the north side of the residential area, indicated by post-holes and `in-situ` wooden posts, indicating a well organized spatial arrangement. Previous studies report an uneven distribution of prestige goods across burials without agreement about whether this uneven distribution hints at vertical social differences. For example, @Cheng2008 interpreted the unequal distribution of glass beads, especially the gold-foil beads between burials, as evidence for hierarchy, indicating a stratified society. However, @Hsieh2012 suggested a relatively egalitarian structure based on comparative analysis of the frequencies of all burial goods. She found that the burials with high value burial goods were usually associated with elders, which might indicate achieved, rather than inherited, status. One important limitation of these previous studies is that they did not use analytical units suitable for comparing pre-European social organization with post-European social organization. Here, we adopt a new chronological framework for the burials to test if network configurations differ from the pre-European period to the post-European period. The discord over the discussion of Kiwulan social organization could be associated with chronological differences that has not yet been well studied.

```{r burial-location, include=FALSE, cache=TRUE, cache.extra=tools::md5sum(here::here("001-data-tidy.R"))}
suppressWarnings(source(here::here("analysis", "paper", "002-burial-location.R")))
```

```{r burial-map, fig.cap = "Map illustrating the location of burials by periods at the central excavation area of Kiwulan (each square is 4 X 4 meters). The gray dots are post holes.", fig.width = image_width_2_col_inch}
KWL_burial_map
```

@@

```{r bergm-pre, include=FALSE, cache=TRUE, cache.extra=tools::md5sum(here::here("003-burials-pre-network-modelling.R"))}
suppressWarnings(source(here::here("analysis", "paper", "003-burials-pre-network-modelling.R")))
```

```{r bergm-post, include=FALSE, cache=TRUE, cache.extra=tools::md5sum(here::here("004-burials-post-network-modelling.R"))}
suppressWarnings(source(here::here("analysis", "paper", "004-burials-post-network-modelling.R")))
```

## Hypothesis and construction of networks

@@

<!-- burial data and the definition of tie formation  --> 
We used burial goods in common to construct ties for representing a flow of good between burials. Trade beads are commonly found across burials with substantial differences in quantities, so we described each burial as having one of four levels, high, upper-middle, lower-middle, and low, according to their distributions across all burials. Gold-foil beads are in levels of high (> 3), upper-middle (= 3), lower-middle (= 2), and low (= 1); carnelian beads are in levels of high (> 6), upper-middle (= 4-6), lower-middle (= 3), and low (= 1-2); glass beads are in levels of high (> 6), upper-middle (= 3-6), lower-middle (= 2), and low (= 1). That means if burial 1 and burial 2 both have high quantities of agate beads, then there will be a tie connecting them. For less frequent prestige goods, including imported porcelains, gold foils, and fish-shaped ornaments, we linked two burials when they possess each type of goods in common (i.e. presence or absence). Node attributes here include osteological data, such as age and sex, and cultural data, such as ritual pottery, and burial value index. Ritual pottery was identified as locally made ceramics placed above graves that suggests funeral feasting according to historical records. The burial value index is an important attribute for economical inequality that we assigned the burials into three levels, high (>= 29, account for 10%), above average (= 12-28), and below average (< 12), as an index of wealth. The number is the sum of values of each type of prestige good, which is calculate by the total number of a prestige item from burial context over the number of prestige item in a burial. Since burials tend to have multiple prestige goods in common, the network ties are weighted instead of binary data (where the value 1 represents a tie and the value 0 otherwise) [@Snijders2011]. For example, if two burials have both low quantity of glass beads and porcelain in common, the tie is given a value of 2. Our networks are non-directed, which means ties have no orientation forming the relationship between actors to indicate a mutual relationship.

@@

```{r make-network-diagrams, include=FALSE, cache=TRUE, cache.extra=tools::md5sum(here::here("005-network-diagrams-two-phases.R"))}
suppressWarnings(source(here::here("analysis", "paper", "005-network-diagrams-two-phases.R")))
```

<!--posterior stats for pre-E-->
The networks constructed from observed burial data show that the network after the European presence has higher node degrees in general (Figure \@ref(fig:network-diagrams)). We compared estimated statistics from the posterior distributions to examine their differences in structure of the simulated networks (Table XX). For the pre-European network simulations, two nodal covariates, age-homophily and ritual-homophily, have significant effects on the formation of relations between burials. The age homophily shows a negative effect that indicates burials of the same age tend to not share similar prestige goods, while the ritual-homophily demonstrates the positive effect that means burials with ritual pottery tend to form relations. Despite positive values for some covariates, such as sex-homophily and burial value-homophily, they do not show a significant tendency to form relations duo to value of zero in the confidence intervals. Similarly, the dyadic covariate, physical distance, shows positive effects but not significantly, indicating that physical proximity does not affect relations. For the endogenous network effects, transitivity (gwesp) present significant positive effect, while popularity (gwdegree) demonstrates negative effect. The positive value for transitivity suggests a tendency of burials to be clustered in closed transitive structures, indicative of the presence of multiple corporate groups sharing burial goods in common. The negative popularity shows there is a tendency toward decentralization that reflects most burials have similar number of ties without any prominent ones. This might imply that individuals have equal access to trade goods in terms of the flow of goods.

```{r network-diagrams, fig.cap = "A: Burial network before the European arrival, B: Burial network after the European arrival. The size of the nodes is proportional to the node degrees that means the number of connections to a node. The thickness of tie represents the number of goods in common between nodes.", fig.width = image_width_3_col_inch}
diagrams_two_phases
```
